A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-07-08 rigid_bodies.tex
Some small changes

\subsection{Kinematics}

@@

As recommended in Ref.~\onlinecite{Martyna1992}, the inertial parameters can be calculated from a characteristic time scale $t_d$ of the system by $Q_1 = L k_B T t_d^2$ and, for $j > 1$, $Q_j = k_B T t_d^2$. Analysis of the probability distribution generated by the equations of motion in Eq.~\ref{eq:nhc_system} can be done via the generalized phase-space analysis of Tuckerman and coworkers.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated. For a system without external forces, the extended energy $H$ and the vector quantity $\vt P e^{\eta_1} = \vt K$ are integrals of motion, while the center-of-mass position $\vt R$ can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, from the $8N$ individual equations implied in Eqs.~\ref{eq:nhc_q} and \ref{eq:nhc_pi}, $2N$ must be eliminated due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$. By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if $L = 6N$.

@@

Considering the analytical solutions described above, it is possible to split the NHC operator into $2M+1$ factors by applying the Trotter-Suzuki formula in a recursive way. The splitting we employ here is
\begin{align*}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = &\prod_{j=M}^1 \exp \left[ \frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right] \\
\times \exp \Bigg[ -\frac{\Delta t}{2} &\alpha_1 \sum_{i=1}^N \left(\tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j} \Bigg] \\
\times &\prod_{j=1}^M \exp \left[\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].\\
\end{align*}

@@

In the microcanonical case, each term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator. As a consequence, the whole propagator preserves the phase-space volume. Recall that the inverse statement is not necessarily true. It is now widely recognized that the symplectic property is important for numerical stability,\cite{Skeel1997} although it is restricted to Hamiltonian systems. For non-Hamiltonian equations of motion, the phase space volume is no longer conserved. Nevertheless, there exists an invariant measure on the phase space, and for the system in Eq.~\ref{eq:nhc_system}, it is given by
\begin{align*}
%\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge \prod_{i=1}^{N}[\vt {r}_i \wedge\vt {p}_i\wedge \vt {q}_i\wedge \vt {\pi}_i ] \wedge \prod_{j=1}^{M}[ \eta_j\wedge p_{\eta_j} ].
\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge {\vt r} \wedge {\vt p} \wedge {\vt q} \wedge {\vt \pi} \wedge {\vt \eta} \wedge {\vt p_\eta}
\end{align*}

@@

We now derive an expression for calculating the pressure of a rigid body system under PBC. For the sake of simplicity, we consider that the interaction potential is pairwise additive with respect to the atoms that form the rigid bodies. This enables us to follow the approach of Louwerse and Baerends.\cite{Louwerse2006} If multibody interactions are present, the approach of Thompson \textit{et al}.\cite{Thompson2009} is recommended.

The potential energy per unit cell of a system with $N$ distinct rigid bodies (indexed from $1$ to $N$) and periodic boundary conditions is
\begin{equation}
\label{eq:U_pairwise_pbc}
U({\vt r}^N, {\vt q}^N) = \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} u({\vt r}_{ij},{\vt q}_i,{\vt q}_j),
\end{equation}
where ${\vt r}_{ij} = {\vt r}_i - {\vt r}_j$ and
\begin{equation}
\label{eq:u_rigid_body_pair}
u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{{\vt R}_i}_k - {{\vt R}_j}_l\| \right).
\end{equation}

In the equation above, $\upsilon(r)$ is the interaction potential between two atoms separated by a distance $r$. The dependence of this potential on the types of these two atoms is omitted for simplicity. The difference ${{\vt R}_i}_k - {{\vt R}_j}_l$ is given by
\[
{{\vt R}_i}_k - {{\vt R}_j}_l = {\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l.
\]

The system pressure, considering a unit cell with constant volume $V$, is\cite{Louwerse2006,Tuckerman2010}
\[
P = \frac{(3N-3) k_B T}{3V} - \left\langle \diff{U}{V} \right\rangle.
\]
The first term in the expression above accounts for the ideal gas contribution. Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
\[
%\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right. \\ \left. \diff{u}{{\vt q}_i} \cdot \diff{{\vt q}_i}{L} + \diff{u}{{\vt q}_j} \cdot \diff{{\vt q}_j}{L}\right).
\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{{\vt r}_{ij}}{L} \diff{u}{{\vt r}_{ij}} + \diff{{\vt q}_i}{L} \diff{u}{{\vt q}_i} + \diff{{\vt q}_j}{L} \diff{u}{{\vt q}_j} \right).
\]

Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = {\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$. Therefore,
\[
\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \tr{\vt r}_{ij} \diff{u}{{\vt r}_{ij}}.
\]

>>>>

\subsection{Kinematics}
\label{sec:kinematics}

@@

As recommended in Ref.~\onlinecite{Martyna1992}, the inertial parameters can be calculated from a characteristic time scale $t_d$ of the system by $Q_1 = L k_B T t_d^2$ and, for $j > 1$, $Q_j = k_B T t_d^2$. Analysis of the probability distribution generated by the equations of motion in Eq.~\ref{eq:nhc_system} can be done via the generalized phase-space analysis of Tuckerman and coworkers.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated. For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, from the $8N$ individual equations implied in Eqs.~\ref{eq:nhc_q} and \ref{eq:nhc_pi}, $2N$ must be eliminated due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$. By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if $L = 6N$.

@@

Considering the analytical solutions described above, it is possible to split the NHC operator into $2M+1$ factors by applying the Trotter-Suzuki formula in a recursive way. The splitting we employ here, noting that the product signs run in reverse orders, is
\begin{align*}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = &\prod_{j=M}^1 \exp \left[ \frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right] \\
\times \exp \Bigg[ -\frac{\Delta t}{2} &\alpha_1 \sum_{i=1}^N \left(\tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j} \Bigg] \\
\times &\prod_{j=1}^M \exp \left[\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].\\
\end{align*}

@@

In the microcanonical case, each term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator. As a consequence, the whole propagator preserves the phase-space volume. Recall that the inverse statement is not necessarily true. It is now widely recognized that the symplectic property is important for numerical stability,\cite{Skeel1997} although it is restricted to Hamiltonian systems. For non-Hamiltonian equations of motion, the phase space volume is no longer conserved. Nevertheless, there exists an invariant measure on the phase space, and for the system in Eq.~\ref{eq:nhc_system}, it is given by
\begin{align*}
\bar{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge {\vt r} \wedge {\vt p} \wedge {\vt q} \wedge {\vt \pi} \wedge {\vt \eta} \wedge {\vt p_\eta}
\end{align*}

@@

The usual way of computing pressure in molecular dynamics is based on.....

We now derive an expression for calculating the pressure of a rigid body system under periodic boundary conditions. For the sake of simplicity, we consider that the interaction potential is pairwise additive with respect to the atoms that form the rigid bodies. This enables us to follow the approach of Louwerse and Baerends.\cite{Louwerse2006} If multibody interactions are present, the approach of Thompson \textit{et al}.\cite{Thompson2009} is recommended. For a pairwise additive model, the potential energy of the $N$ rigid bodies whose centers of mass lie in a simulation box with periodic boundary conditions is
\begin{equation}
\label{eq:U_pairwise_pbc}
U({\vt r}, {\vt q}) = \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} u({\vt r}_{ij},{\vt q}_i,{\vt q}_j),
\end{equation}
where ${\vt r}_{ij} = {\vt r}_i - {\vt r}_j$ and
\begin{equation}
\label{eq:u_rigid_body_pair}
u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{{\vt R}_i}_k - {{\vt R}_j}_l\| \right).
\end{equation}

\begin{equation}
\label{eq:u_rigid_body_pair}
u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l\| \right).
\end{equation}

In the equation above, ${{\vt R}_i}_k$ is the position of the $k$-th particle of body $i$ and $\upsilon(r)$ is the interaction potential between two atoms separated by a distance $r$. The dependence of this potential on the types of these two atoms is omitted for simplicity. From Sec.~\ref{sec:kinematics}, one can deduce that

The difference ${{\vt R}_i}_k - {{\vt R}_j}_l$ is given by
\[
{{\vt R}_i}_k - {{\vt R}_j}_l = {\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l.
\]

The system pressure, considering a unit cell with constant volume $V$, is\cite{Louwerse2006,Tuckerman2010}
\[
P = \frac{(3N-3) k_B T}{3V} - \left\langle \diff{U}{V} \right\rangle.
\]

The first term in the expression above accounts for the ideal gas contribution. Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
\[
%\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right. \\ \left. \diff{u}{{\vt q}_i} \cdot \diff{{\vt q}_i}{L} + \diff{u}{{\vt q}_j} \cdot \diff{{\vt q}_j}{L}\right).
\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{{\vt r}_{ij}}{L} \diff{u}{{\vt r}_{ij}} + \diff{{\vt q}_i}{L} \diff{u}{{\vt q}_i} + \diff{{\vt q}_j}{L} \diff{u}{{\vt q}_j} \right).
\]

Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = \tr{\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$. Therefore,
\[
\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \tr{\vt r}_{ij} \diff{u}{{\vt r}_{ij}}.
\]
